MassWateR: Improving quality control, analysis, and sharing of water quality data

### 2023-09-19 revision/review-response.qmd

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib

@@

Line 26 includes a link to a description of the US Clean Water Act ([33 USC § 1251 et seq.](https://uscode.house.gov/view.xhtml?path=/prelim@title33/chapter26&edition=prelim), note that there is no citation provided in the references because this is legislative code).

Line 29 includes a citation to the European Water Framework Directive [Directive 2000/60/EC of the European Parliament and of the Council of 23 October 2000, @pollard1998]

Line 40 includes the following citations: @kinzelman2003, @Wagner06, @zhang2006.

Lines 521-537 include the following citations: @Meyer22, @whitehead2007, @list2017, @Wickham16

Additional citations were added later in the discussion: @DeCicco22, @Wagner06

@@

4.  Figure 1 could best be presented as a flow/block diagram.

**Response**: Figure 1 has been reconfigured as a flow diagram, although emphasis remains in the text that a user can engage with the package using one to any of the steps in the diagram.

5.  For Table 4, it would be important to provide all the parameters tested as a supplementary file to support the submission. Please ensure that the full details of the algorithms designed are provided, and where there are restrictions, it should otherwise be stated.

**Response**: A supplement file was created to accompany the main manuscript, including tables for the parameters used by MassWateR and the sample files included with the package for the data quality objectives for accuracy, frequency, and completeness. Also note that the specific QC checks used by the package were previously described in the main text (e.g., lab spikes, lab blanks, etc.).

### 2023-09-19 manu-draft.qmd
more revisions, added supplement

After all file format checks are completed and errors fixed, a standard approach for using *MassWateR* is to import all required files and save them as a list of named data frame objects that can be used by nearly all the package functions. This prevents the need to identify which input datasets are needed for each function, although the latter approach could be used because arguments for individual input files are also provided in all functions. In the latter case, a path or data object can be used as input for each file. For the former approach, the beginning of a script for using the package could appear as follows. Example files included with the package are imported, whereas a user will specify paths to their own files.

@@

A final note about the data inputs is that many types of water quality measurements can be included for analysis, although the package is currently limited to working with discrete samples as compared to data from continuous monitoring equipment. The `paramsMWR` dataset included with *MassWateR* provides a lookup table of the parameters that can be used with the package. On data import, this dataset is referenced to ensure that only relevant parameters are included and that appropriate units of measurement are provided. The dataset includes `r nrow(paramsMWR)` different parameters, each with multiple valid units of measurement. Additionally, only one unit of measurement is allowed per parameter, which was an intentional design so that tedious functions for converting between dozens of units of measurement did not need to be created during package development. It is not an unreasonable expectation for users to provide only one unit of measurement per parameter. Laboratories typically have a standard reporting format based on the same methodology or instrument used to measure water quality parameters. Both the number of parameters and acceptable units can easily be extended and those herein were chosen simply to limit the development of *MassWateR* to a manageable scope.

@@

The quality control functions in *MassWateR* are designed to create a single report that compares the water quality data in the results file (`resdat`) to data quality objectives in the accuracy (`accdat`) and frequency and completeness (`frecomdat`) files. In general, the QC checks for accuracy evaluate if laboratory and field duplicates, blanks, or spikes are within acceptable ranges. The QC checks for frequency and completeness can be used to evaluate if a sufficient number of records in the results file satisfy the accuracy checks and that sufficient QC data have been collected. The values in the `accdat` and `frecomdat` inputs are collectively described as data quality objectives (DQOs), such that the QC samples in `resdat` must satisfy these objectives to be considered accurate and precise data for use in regulatory or other assessments by water quality professionals.

@@

To ensure that *MassWateR* is a known resource for potential users and to encourage its use for QC reporting, analysis, and data submission, a community of practice was established during package development and following its initial release on CRAN (January, 2023). This included several beta testing and training workshops to gather feedback on anticipated data analysis workflows and to educate potential users on appropriate use of the package. Many users have not previously been exposed to R for data analysis and a substantial portion of the trainings included an [introduction](https://massbays-tech.github.io/intro-to-r/) to R, as well as *MassWateR*. Emphasis was placed on simple use of the core functions, as opposed to developing custom workflows that combined core R functions with *MassWateR*. However, users were encouraged to learn how to extend the use of the package by leveraging additional R packages, such as *ggplot2* to create additional visualizations for modifying the existing analysis plots in *MassWateR*. For example, a [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html) was included on the package website to demonstrate how these plots can be modified for custom output using *ggplot2*. Further, a [community forum](https://massbays.discourse.group/c/masswater-r-tools/5) was created as a resource for users to post questions about the package and for others to view the discussion if similar issues were encountered. This approach followed the model used by other popular web forums (e.g., StackOverflow) for troubleshooting software issues by minimizing duplication of topics through sharing solutions in a public forum.

@@

Additional future work to improve the functionality of *MassWateR* is also expected as the user base increases and the functions mature with additional bug fixes or minor enhancements. Specifically, the inclusion of historical data for some of the analysis functions could provide additional context on status and trends for monitoring data at specific locations. This enhancement would require the extraction of existing data included in the WQP, which would not be significantly challenging given the robust web retrieval tools already available. Existing R packages leverage these tools (e.g., *dataRetrieval*) and a similar approach could be used by *MassWateR*. A second and more challenging enhancement would be the ability to work with continuous monitoring data collected at high temporal resolution with equipment deployed *in situ*. These data present additional challenges not encountered with routine samples collected at longer time intervals, including increased data volume and additional QC needs [@Wagner06; @Horsburgh15]. For the latter, automated tools are needed for detecting and handling QC issues common with monitoring equipment deployed in the field for long duration, such as instrument drift, biofouling, or missing data. More complex methods for detecting outliers in continuous monitoring data beyond the existing tools in *MassWateR* will also be needed [@Leigh19].

>>>>

After all file format checks are completed and errors fixed, a standard approach for using *MassWateR* is to import all required files and save them as a list of named data frame objects that can be used by nearly all the package functions. This prevents the need to identify which input datasets are needed for each function, although the latter approach could be used because arguments for individual input files are also provided in all functions. In the latter case, a path or data object can be used as input for each file. For the former approach, the beginning of a script for using the package could appear as follows. Example files included with the package are imported (see the supplement for the example data quality objective files for accuracy, frequency and completeness), whereas a user will specify paths to their own files.

@@

A final note about the data inputs is that many types of water quality measurements can be included for analysis, although the package is currently limited to working with discrete samples as compared to data from continuous monitoring equipment. The `paramsMWR` dataset included with *MassWateR* provides a lookup table of the parameters that can be used with the package (see supplement). On data import, this dataset is referenced to ensure that only relevant parameters are included and that appropriate units of measurement are provided. The dataset includes `r nrow(paramsMWR)` different parameters, each with multiple valid units of measurement. Additionally, only one unit of measurement is allowed per parameter, which was an intentional design so that tedious functions for converting between dozens of units of measurement did not need to be created during package development. It is not an unreasonable expectation for users to provide only one unit of measurement per parameter. Laboratories typically have a standard reporting format based on the same methodology or instrument used to measure water quality parameters. Both the number of parameters and acceptable units can easily be extended and those herein were chosen simply to limit the development of *MassWateR* to a manageable scope.

@@

The quality control functions in *MassWateR* are designed to create a single report that compares the water quality data in the results file (`resdat`) to data quality objectives in the accuracy (`accdat`, see the supplement for the sample file included with the package) and frequency and completeness (`frecomdat`, see the supplement for the sample file included with the package) files. In general, the QC checks for accuracy evaluate if laboratory and field duplicates, blanks, or spikes are within acceptable ranges. The QC checks for frequency and completeness can be used to evaluate if a sufficient number of records in the results file satisfy the accuracy checks and that sufficient QC data have been collected. The values in the `accdat` and `frecomdat` inputs are collectively described as data quality objectives (DQOs), such that the QC samples in `resdat` must satisfy these objectives to be considered accurate and precise data for use in regulatory or other assessments by water quality professionals.

@@

To ensure that *MassWateR* is a known resource for potential users and to encourage its use for QC reporting, analysis, and data submission, a community of practice [e.g., @Meyer22] was established during package development and following its initial release on CRAN (January, 2023). This included several beta testing and training workshops to gather feedback on anticipated data analysis workflows and to educate potential users on appropriate use of the package [@whitehead2007; @list2017]. Many users have not previously been exposed to R for data analysis and a substantial portion of the trainings included an [introduction](https://massbays-tech.github.io/intro-to-r/) to R, as well as *MassWateR*. Emphasis was placed on simple use of the core functions, as opposed to developing custom workflows that combined core R functions with *MassWateR*. However, users were encouraged to learn how to extend the use of the package by leveraging additional R packages, such as *ggplot2* [@Wickham16] to create additional visualizations for modifying the existing analysis plots in *MassWateR*. For example, a [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html) was included on the package website to demonstrate how these plots can be modified for custom output using *ggplot2*. Further, a [community forum](https://massbays.discourse.group/c/masswater-r-tools/5) was created as a resource for users to post questions about the package and for others to view the discussion if similar issues were encountered. This approach followed the model used by other popular web forums (e.g., \[StackOverflow\](https://stackoverflow.com/)) for troubleshooting software issues by minimizing duplication of topics through sharing solutions in a public forum.

@@

Additional future work to improve the functionality of *MassWateR* is also expected as the user base increases and the functions mature with additional bug fixes or minor enhancements. Specifically, the inclusion of historical data for some of the analysis functions could provide additional context on status and trends for monitoring data at specific locations. This enhancement would require the extraction of existing data included in the WQP, which would not be significantly challenging given the robust web retrieval tools already available. Existing R packages leverage these tools [e.g., *dataRetrieval,* @DeCicco22] and a similar approach could be used by *MassWateR*. A second and more challenging enhancement would be the ability to work with continuous monitoring data collected at high temporal resolution with equipment deployed *in situ* [e.g., @Wagner06]. These data present additional challenges not encountered with routine samples collected at longer time intervals, including increased data volume and additional QC needs [@Wagner06; @Horsburgh15]. For the latter, automated tools are needed for detecting and handling QC issues common with monitoring equipment deployed in the field for long duration, such as instrument drift, biofouling, or missing data. More complex methods for detecting outliers in continuous monitoring data beyond the existing tools in *MassWateR* will also be needed [@Leigh19].
